National Longitudinal Evidence for Growth in Subjective Wellbeing from Spiritual Beliefs

### 2020-11-25 Manuscript_Bayesian.Rmd
Add files via upload
bayesian replication document

---
title             : "National Longitudinal Evidence for Growth in Subjective Wellbeing from Spiritual Beliefs"
shorttitle        : "Longitudinal Wellbeing from Spiritual Beliefs"
author: 
  - name          : "Benjamin Highland"
    affiliation   : "1"
    corresponding : yes    # Define only one corresponding author
    address       : "520 Power Plant Circle Apartment 105 Winston-Salem, NC 27101 USA"
    email         : "benhighland6420@gmail.com"
  - name          : "Everett L. Worthington"
    affiliation   : "2"
    institution   : "Department of Psychology Virginia Commonwealth University"
  - name          : "Don E Davis"
    affiliation   : "3"
    institution   : "Department of Psychology Georgia State University"
  - name          : "Chris G. Sibley"
    affiliation   : "4"
    institution   : "School of Psychology University of Auckland"
  - name          : "Joseph A. Bulbulia"
    affiliation   : "5"
    institution   : "School of Psychology Victoria University"

@@

abstract: |

  Previous research finds an association between spirituality and subjective well-being. However, the widespread use of tautological spirituality scales, poorly defined concepts of spirituality, and heavy reliance on cross-sectional samples cast doubts. Here, we leverage nine waves of panel data from a nationally diverse longitudinal study to systematically test whether having spiritual beliefs leads to growth in personal well-being and life satisfaction ($N$ = 20979, New Zealand, 2010-2018). Contrary to previous research, we find that belief in a spirit or life force predicts lower personal well-being and life-satisfaction. However, in support of previous speculation, beliefs in a spirit or life force predict increasing personal well-being and life satisfaction over time relative to disbelief. These finding are robust after to known demographic influences, and intriguingly, hold among those who believe in a God but disbelieve in a Spirit or Life Force. The recent growth in spiritual beliefs and decline in traditional religion across many industrial societies motivates further causal investigations of the mechanisms by which spiritual beliefs lead to growth in subjective well-being. 
  <!-- https://tinyurl.com/ybremelq -->

keywords          : "Belief, God, Health, Longitudinal, Panel, Religion, Spirit, Spirituality, Well-being"
wordcount         : "5415 words in text body, 1155 words in reference section"
bibliography      : ["b-references.bib"]

floatsintext      : no
figurelist        : no
figsintext        : no
tablelist         : no
footnotelist      : no
linenumbers       : no
mask              : no
draft             : no

documentclass     : "apa6"
classoption       : "man"
output            : papaja::apa6_pdf
extra_dependencies: ["rotating", "booktabs", "graphicx","longtable","graphicx"]
header-includes:
   - \usepackage{tcolorbox}
   - \usepackage{paralist}
   - \usepackage{multicol}
   - \usepackage{graphics}
   - \usepackage{rotating}
   - \usepackage{longtable}
---
```{r libraries}
library("tidyverse")
library("papaja")
library("here")
```

@@

<!-- In spirituality/psychological wellbeing research, subjective in spirituality health research, psychological well-being is typically decomposed into two categories: emotional well-being (i.e., the presence or absence of positive feelings about life) and functional well-being (i.e., psychological and social well-being) [@Keyes2002-bb]. -->

First, previous research has frequently operationalised spirituality using measures of well-being. For example, The *Daily Spiritual Experience Scale*, which is widely used in spiritual mental health research, includes items such as "I feel thankful for my blessings" and "I feel deep inner peace or harmony" to measure spirituality [@Underwood2002-hg]. As Koenig observes, it is unsurprising that spirituality predict well-being when the same items are used to measure both constructs. Koenig aptly describes this approach as "tautological" [@Koenig2008-lv]. Indeed, recent systematic-review of spirituality/mental health research found that nearly 45% of previous studies employ tautological scales [@Garssen2016-kb]. The first challenge, then, is to define measures of spirituality that are not well-being measures dressed up in other words.

@@

Here, we addresses the three challenges of previous in the following ways: First, to avoid a tautological elision of "spirituality" and "well-being" we adopt a non-affective measure of spirituality as a "believe in a Spirit or Life Force." On the other side, to measure subjective well-being, we use a well-validated scale for personal well-being and a well-validated scale for Life-Satisfaction (see Method). Second, to disentangle spiritual beliefs from traditional religious beliefs and skeptical beliefs we adopt a four-level categorical indicator combining the binary indicator of belief in a Spirit or Life Force with the binary indicator "Do you believe in some form of a spirit or life force?" The levels of this four-level categorical indicator for spiritual, religious, and skeptical beliefs are as follows: (1) Skeptics: those who neither believe in a God nor believe in a Spirit or Life force; (2) Skeptics about God who nevertheless believe in a Spirit or Life Force; (3) Those who believe in a God but do not believe in a Spirit or Life force; (4) Those who who believe in a both God and a Spirit or Life Force. Clearly such an approach cannot simultaneously address all interests in previous spirituality/well-being research. For example we do not combine our cognitive measure of belief with behavioral measures or attitudinal measures. However, this cost to scope is paid for with benefits to precision. The spiritual indicators measure cognitive states in which distinct varieties of skeptical beliefs, traditional religious beliefs, and non-traditional spirit beliefs are straightforwardly disentangled, and measured within people. Third, to assess whether spirit beliefs lead to growth, stability, or decline in psychological well-being, we leverage longitudinal responses from a nationally diverse probability sample year over an nine--year period with nine measurement points (2010 to 2018 New Zealand Attitudes and Values Study, NZAVS).

@@

Table 1 about here:
\[
\begin{table}[ ht ] 
\centering 
\caption{}\label{}
\scalebox{.6}{
\begin{tabular}{ l c c c c c c c c c }
\toprule
 &   \multicolumn{ 9 }{c}{ Wave }\\ 
  & 2010 & 2011 & 2012 & 2013 & 2014 & 2015 & 2016 & 2017 & 2018 \\ 
 & n = \emph{missing} & n = \emph{missing} & n = 8547 & n = 13472 & n = \emph{missing} & n = 11687 & n = 16048 & n = 14174 & n = 14883 \\ 
 \midrule
Age &   &   &   &   &   &   &   &   &  \\ 
NZdep &   &   &   &   &   &   &   &   &  \\ 
Education &   &   &   &   &   &   &   &   &  \\ 
Employed &   &   &   &   &   &   &   &   &  \\ 
\hspace{6pt}    0 &  (\%) &  (\%) & 2094 (24.5\%) & 2955 (21.9\%) &  (\%) & 2525 (21.6\%) & 3486 (21.7\%) & 3223 (22.7\%) & 3404 (22.9\%)\\ 
\hspace{6pt}    1 &  (\%) &  (\%) & 6453 (75.5\%) & 10517 (78.1\%) &  (\%) & 9162 (78.4\%) & 12562 (78.3\%) & 10951 (77.3\%) & 11479 (77.1\%)\\ 
EthnicCategories &   &   &   &   &   &   &   &   &  \\ 
\hspace{6pt}    Euro &  (\%) &  (\%) & 6731 (78.8\%) & 11024 (81.8\%) &  (\%) & 9671 (82.8\%) & 13389 (83.4\%) & 11814 (83.3\%) & 12427 (83.5\%)\\ 
\hspace{6pt}    Maori &  (\%) &  (\%) & 1188 (13.9\%) & 1608 (11.9\%) &  (\%) & 1341 (11.5\%) & 1749 (10.9\%) & 1620 (11.4\%) & 1677 (11.3\%)\\ 
\hspace{6pt}    Pacific &  (\%) &  (\%) & 284 (3.3\%) & 339 (2.5\%) &  (\%) & 277 (2.4\%) & 331 (2.1\%) & 243 (1.7\%) & 261 (1.8\%)\\ 
\hspace{6pt}    Asian &  (\%) &  (\%) & 344 (4\%) & 501 (3.7\%) &  (\%) & 398 (3.4\%) & 579 (3.6\%) & 497 (3.5\%) & 518 (3.5\%)\\ 
Male &   &   &   &   &   &   &   &   &  \\ 
\hspace{6pt}    0 &  (\%) &  (\%) & 5302 (62\%) & 8457 (62.8\%) &  (\%) & 7325 (62.7\%) & 10148 (63.2\%) & 9011 (63.6\%) & 9433 (63.4\%)\\ 
\hspace{6pt}    1 &  (\%) &  (\%) & 3245 (38\%) & 5015 (37.2\%) &  (\%) & 4362 (37.3\%) & 5900 (36.8\%) & 5163 (36.4\%) & 5450 (36.6\%)\\ 
Partner &   &   &   &   &   &   &   &   &  \\ 
\hspace{6pt}    0 &  (\%) &  (\%) & 2408 (28.2\%) & 3546 (26.3\%) &  (\%) & 2815 (24.1\%) & 3773 (23.5\%) & 3330 (23.5\%) & 3556 (23.9\%)\\ 
\hspace{6pt}    1 &  (\%) &  (\%) & 6139 (71.8\%) & 9926 (73.7\%) &  (\%) & 8872 (75.9\%) & 12275 (76.5\%) & 10844 (76.5\%) & 11327 (76.1\%)\\ 
Pol Orient &   &   &   &   &   &   &   &   &  \\ 
Urban &   &   &   &   &   &   &   &   &  \\ 
\hspace{6pt}    Not\_Urban &  (\%) &  (\%) & 2865 (33.5\%) & 4278 (31.8\%) &  (\%) & 3951 (33.8\%) & 5538 (34.5\%) & 2610 (18.4\%) & 2661 (17.9\%)\\ 
\hspace{6pt}    Urban &  (\%) &  (\%) & 5682 (66.5\%) & 9194 (68.2\%) &  (\%) & 7736 (66.2\%) & 10510 (65.5\%) & 11564 (81.6\%) & 12222 (82.1\%)\\ 
Beliefs &   &   &   &   &   &   &   &   &  \\ 
\hspace{6pt}    \_Skeptic\_ &  (\%) &  (\%) & 2048 (24\%) & 3286 (24.4\%) &  (\%) & 2951 (25.3\%) & 4116 (25.6\%) & 3916 (27.6\%) & 4120 (27.7\%)\\ 
\hspace{6pt}    \_SpiritExcludesGod\_ &  (\%) &  (\%) & 2295 (26.9\%) & 3735 (27.7\%) &  (\%) & 3169 (27.1\%) & 4501 (28\%) & 3918 (27.6\%) & 3970 (26.7\%)\\ 
\hspace{6pt}    GodAndSpirit &  (\%) &  (\%) & 3707 (43.4\%) & 5499 (40.8\%) &  (\%) & 5119 (43.8\%) & 6453 (40.2\%) & 5464 (38.5\%) & 6210 (41.7\%)\\ 
\hspace{6pt}    GodExcludesSpirit &  (\%) &  (\%) & 497 (5.8\%) & 952 (7.1\%) &  (\%) & 448 (3.8\%) & 978 (6.1\%) & 876 (6.2\%) & 583 (3.9\%)\\ 
LIFESAT &   &   &   &   &   &   &   &   &  \\ 
PWI &   &   &   &   &   &   &   &   &  \\ 
\bottomrule
\end{tabular}
}
\end{table}
\]

## Demographic Indicators
We included standard demographic indicators used in New Zealand Attitudes and Values Research: [NZAVS modelling standards](https://cdn.auckland.ac.nz/assets/psych/about/our-research/nzavs/NZAVSTechnicalDocuments/NZAVS-Technical-Documents-e11-Default-Statistical-Standards.pdf)

@@

## Statistical Analyses
Statistical analysis was performed using R version 4.0.2 R version 3.6.1 (2020-06-22) 2019-07-05). We analyze data from participants who responded to the NZAVS survey at least three times between 2010 (Time 2) and 2018 (Time 10), resulting in a sample of  21,705 (N = 20,979 with complete responses).  To model growth in subjective well-being we use generalized linear mixed models. In our models, we account for the non-linear effects of time within participants by including the interaction of years (a continuous variable) with Beliefs (a four-level categorical indicator). The baseline for this indicator is Skepticism about a Spirit/Life Forces and Skepticism about a God. To handle the dependencies and heterogeneity introduced from the repeated measures, we include individual ID as an effect modelled as random.  We centered and scaled the education, deprivation, and political conservativism. We centred age was centred at and put into decade-long intervals. 
The model equation(s) are described in Equation 1.

## Equation 1
\[
\begin{aligned}
\operatorname{{\small LIFESAT/PWI}}  &\sim N \left(\mu,\sigma^2 \right) \\ \mu &=\alpha_{j[i]} + \beta_{1}(\operatorname{Years}) + \beta_{2}(\operatorname{Beliefs}_{\operatorname{SpiritExcludesGod}})\ + \\
&\quad \beta_{3}(\operatorname{Beliefs}_{\operatorname{GodAndSpirit}}) + \beta_{4}(\operatorname{Beliefs}_{\operatorname{GodExcludesSpirit}}) + \beta_{5}(\operatorname{Age.10yrs.C})\ + \\
&\quad \beta_{6}(\operatorname{Deprivation.S}) + \beta_{7}(\operatorname{Edu.S}) + \beta_{8}(\operatorname{Employed})\ + \\
&\quad \beta_{9}(\operatorname{EthnicCats}_{\operatorname{Maori}}) + \beta_{10}(\operatorname{EthnicCats}_{\operatorname{Pacific}}) + \beta_{11}(\operatorname{EthnicCats}_{\operatorname{Asian}})\ + \\
&\quad \beta_{12}(\operatorname{Male}_{\operatorname{1}}) + \beta_{13}(\operatorname{Partner}_{\operatorname{Has\ }}) + \beta_{14}(\operatorname{Pol.Orient.S})\ + \\
&\quad \beta_{15}(\operatorname{Urban}) + \beta_{16}(\operatorname{Years} \times \operatorname{Beliefs}_{\operatorname{SpiritExcludesGod}}) + \beta_{17}(\operatorname{Years} \times \operatorname{Beliefs}_{\operatorname{GodAndSpirit}})\ + \\
&\quad \beta_{18}(\operatorname{Years} \times \operatorname{Beliefs}_{\operatorname{GodExcludesSpirit}}) \\ \alpha_{j} &\sim N \left(\mu_{\alpha_{j}},\sigma^2_{\alpha_{j}} \right) , \operatorname{ for  Id }~j~= 1, \dots~J = 20,979~\operatorname{Individuals}
\end{aligned}
\]

### System and packages
The analysis was performed using R version 4.0.2 (2020-06-22). The Platform was x86_64-apple-darwin17.0 (64-bit) Running under: macOS Catalina 10.15.6
We are greatful to the contributors and mantainers of the following packages:
R [Version 4.0.2; @R-base] and the R-packages *Amelia* [Version 1.7.6; @R-Amelia], *citr* [Version 0.3.2; @R-citr], *dplyr* [Version 1.0.2; @R-dplyr], *equatiomatic* [Version 0.1.0.9000; @R-equatiomatic], *forcats* [Version 0.5.0; @R-forcats], *ggeffects* [Version 0.15.0; @R-ggeffects], *gghighlight* [Version 0.3.0; @R-gghighlight], *ggplot2* [Version 3.3.2; @R-ggplot2], *ggsci* [Version 2.9; @R-ggsci], *lme4* [Version 1.1.23; @R-lme4], *Matrix* [Version 1.2.18; @R-Matrix], *papaja* [Version 0.1.0.9997; @R-papaja], *parameters* [Version 0.8.6.1; @R-parameters], *patchwork* [Version 1.0.1; @R-patchwork], *prettycode* [Version 1.1.0; @R-prettycode], *purrr* [Version 0.3.4; @R-purrr], *Rcpp* [Version 1.0.5.1; @R-Rcpp_a; @R-Rcpp_b], *readr* [Version 1.3.1; @R-readr], *report* [Version 0.1.0; @R-report], *see* [Version 0.5.2.1; @R-see], *sjPlot* [Version 2.8.4; @R-sjPlot], *stringr* [Version 1.4.0; @R-stringr], *styler* [Version 1.3.2; @R-styler], *table1* [Version 1.2; @R-table1], *texreg* [Version 1.37.5; @R-texreg], *tibble* [Version 3.0.4; @R-tibble], *tidyr* [Version 1.1.2; @R-tidyr], and *tidyverse* [Version 1.3.0; @R-tidyverse]"

### Missing data
Because missing values can result in biased estimates [@Blackwell2017-oq], we multiply imputed 10 datasets using the AmeliaII package in R[@Honaker2011-yu].We conducted a parallel analysis averaging over the multiply imputed datasets, without any substantive difference to inferences. Additionally the results were robust to filtering of participants of N >3 to N = 9 waves. Additionally, the findings reported below were robust to modeling only the effects of years and beliefs, without adjusting for the standard NZAVS demographic covariates. The supplement describing these analysis is located at: https://github.com/jacopastorius/nzavs_spiritbeliefWellbeing.

@@

\[
\begin{table}
\begin{center}
\scalebox{0.6}{
\begin{tabular}{l c c}
\toprule
 & PWI & Life Satisfaction \\
\midrule
(Intercept)                        & $\mathbf{6.63}^{***}$  & $\mathbf{4.94}^{***}$  \\
                                   & $(0.03)$               & $(0.02)$               \\
Years                              & $0.00$                 & $-0.00$                \\
                                   & $(0.00)$               & $(0.00)$               \\
Beliefs\_SpiritExcludesGod\_       & $\mathbf{-0.14}^{***}$ & $\mathbf{-0.08}^{***}$ \\
                                   & $(0.03)$               & $(0.02)$               \\
BeliefsGodAndSpirit                & $\mathbf{-0.16}^{***}$ & $-0.04$                \\
                                   & $(0.03)$               & $(0.02)$               \\
BeliefsGodExcludesSpirit           & $0.02$                 & $0.07$                 \\
                                   & $(0.05)$               & $(0.04)$               \\
Age.10yrs.C                        & $\mathbf{0.08}^{***}$  & $\mathbf{0.09}^{***}$  \\
                                   & $(0.01)$               & $(0.01)$               \\
Deprivation.S                      & $\mathbf{-0.14}^{***}$ & $\mathbf{-0.07}^{***}$ \\
                                   & $(0.01)$               & $(0.01)$               \\
Edu.S                              & $\mathbf{0.12}^{***}$  & $\mathbf{0.06}^{***}$  \\
                                   & $(0.01)$               & $(0.01)$               \\
EmployedEmployed                   & $\mathbf{0.09}^{***}$  & $\mathbf{0.08}^{***}$  \\
                                   & $(0.01)$               & $(0.01)$               \\
EthnicCatsMaori                    & $\mathbf{-0.30}^{***}$ & $\mathbf{-0.06}^{**}$  \\
                                   & $(0.03)$               & $(0.02)$               \\
EthnicCatsPacific                  & $\mathbf{-0.29}^{***}$ & $\mathbf{-0.08}^{*}$   \\
                                   & $(0.05)$               & $(0.04)$               \\
EthnicCatsAsian                    & $-0.08$                & $-0.04$                \\
                                   & $(0.05)$               & $(0.03)$               \\
Male1                              & $\mathbf{-0.11}^{***}$ & $\mathbf{-0.19}^{***}$ \\
                                   & $(0.02)$               & $(0.01)$               \\
PartnerHas Partner                 & $\mathbf{0.61}^{***}$  & $\mathbf{0.42}^{***}$  \\
                                   & $(0.02)$               & $(0.01)$               \\
Pol.Orient.S                       & $\mathbf{0.02}^{***}$  & $\mathbf{0.01}^{**}$   \\
                                   & $(0.01)$               & $(0.00)$               \\
UrbanUrban                         & $\mathbf{-0.02}^{*}$   & $\mathbf{-0.02}^{*}$   \\
                                   & $(0.01)$               & $(0.01)$               \\
Years:Beliefs\_SpiritExcludesGod\_ & $\mathbf{0.02}^{**}$   & $\mathbf{0.02}^{***}$  \\
                                   & $(0.00)$               & $(0.00)$               \\
Years:BeliefsGodAndSpirit          & $\mathbf{0.02}^{***}$  & $\mathbf{0.01}^{***}$  \\
                                   & $(0.00)$               & $(0.00)$               \\
Years:BeliefsGodExcludesSpirit     & $0.00$                 & $-0.00$                \\
                                   & $(0.01)$               & $(0.01)$               \\
\midrule
AIC                                & $248580.18$            & $200599.16$            \\
BIC                                & $248775.08$            & $200793.96$            \\
Log Likelihood                     & $-124269.09$           & $-100278.58$           \\
Num. obs.                          & $79270$                & $78888$                \\
Num. groups: Id                    & $20979$                & $20972$                \\
Var: Id (Intercept)                & $1.78$                 & $0.88$                 \\
Var: Residual                      & $0.74$                 & $0.42$                 \\
\bottomrule
\multicolumn{3}{l}{\scriptsize{$^{***}p<0.001$; $^{**}p<0.01$; $^{*}p<0.05$}}
\end{tabular}
}
\caption{Regression Models}
\label{table:coefficients}
\end{center}
\end{table}
\]

@@

>>>>

---
title             : "Longitudinal Evidence for Growth in Subjective Wellbeing from Spiritual Belief"
shorttitle        : "Longitudinal Wellbeing from Spiritual Beliefs"
author: 
  - name          : "Benjamin Highland"
    affiliation   : "1"
    corresponding : yes    # Define only one corresponding author
    address       : "520 Power Plant Circle Apartment 105 Winston-Salem, NC 27101 USA"
    email         : "benhighland6420@gmail.com"
  - name          : "Everett L. Worthington"
    affiliation   : "2"
    institution   : "Department of Psychology Virginia Commonwealth University"
  - name          : "Don E Davis"
    affiliation   : "3"
    institution   : "Department of Psychology Georgia State University"
  - name          : "Chris G. Sibley"
    affiliation   : "4"
    institution   : "School of Psychology University of Auckland"
  - name          : "Joseph A. Bulbulia"
    affiliation   : "5"
    institution   : "School of Psychology Victoria University"

@@

abstract: | 
  Previous research finds an association between spirituality and subjective well-being. However, the widespread use of tautological spirituality scales, poorly defined concepts of spirituality, and heavy reliance on cross-sectional samples cast doubts. Here, we leverage nine waves of panel data from a nationally diverse longitudinal study to systematically test whether having spiritual beliefs leads to growth in personal well-being and life satisfaction ($N$ = 20979, New Zealand, 2010-2018). Contrary to previous research, we find that belief in a spirit or life force predicts lower personal well-being and life-satisfaction. However, in support of previous speculation, beliefs in a spirit or life force predict increasing personal well-being and life satisfaction over time relative to disbelief. These finding are robust after to known demographic influences, and intriguingly, hold among those who believe in a God but disbelieve in a Spirit or Life Force. The recent growth in spiritual beliefs and decline in traditional religion across many industrial societies motivates further causal investigations of the mechanisms by which spiritual beliefs lead to growth in subjective well-being. 
  <!-- https://tinyurl.com/ybremelq -->

keywords          : "Belief, God, Health, Longitudinal, Panel, Religion, Spirit, Spirituality, Well-being"
wordcount         : "5415 words in text body, 1155 words in reference section"
bibliography      : ["b-references.bib"]
floatsintext      : no
figurelist        : no
figsintext        : no
tablelist         : no
footnotelist      : no
linenumbers       : no
mask              : no
draft             : no
documentclass     : "apa6"
classoption       : "man"
output            : papaja::apa6_pdf
extra_dependencies: ["rotating", "booktabs", "graphicx","longtable","amsmath","amsfonts","amssymb"]
# header-includes:
#    - \usepackage{tcolorbox}
#    - \usepackage{paralist}
#    - \usepackage{multicol}
#    - \usepackage{graphics}
#    - \usepackage{rotating}
#    - \usepackage{longtable}
#    - \usepackage[amsmath]
---
```{r libraries}
library("tidyverse")
library("papaja")
library("here")
```

@@

First, previous research has frequently operationalised spirituality using measures of well-being. For example, The *Daily Spiritual Experience Scale*, which is widely used in spiritual mental health research, includes items such as "I feel thankful for my blessings" and "I feel deep inner peace or harmony" to measure spirituality [@Underwood2002-hg]. These items are used both to measure spirituality and subjective well-being. As Koenig observes, it is unsurprising that spirituality is associated with well-being when the same items are used to measure both constructs. Koenig aptly describes this problematic approach as "tautological" [@Koenig2008-lv]. How pervasive is this problem? A recent systematic-review of spirituality/mental health research found that nearly 45% of previous studies employ tautological scales [@Garssen2016-kb]. The first challenge, then, is to define measures of spirituality that are not well-being measures dressed up in other words.

@@

Here, we addresses these three challenges of previous research in the following ways: First, to avoid a tautological elision of "spirituality" and "well-being" we adopt a non-affective measure of spirituality as a "believe in a Spirit or Life Force." On the other side, to measure subjective well-being, we use a well-validated scale for personal well-being and a well-validated scale for Life-Satisfaction (see Method). Second, to disentangle spiritual beliefs from traditional religious beliefs and skeptical beliefs we adopt a four-level categorical indicator combining the binary indicator of belief in a Spirit or Life Force with the binary indicator "Do you believe in some form of a spirit or life force?" The levels of this four-level categorical indicator for spiritual, religious, and skeptical beliefs are as follows: (1) Skeptics: those who neither believe in a God nor believe in a Spirit or Life force; (2) Skeptics about God who nevertheless believe in a Spirit or Life Force; (3) Those who believe in a God but do not believe in a Spirit or Life force; (4) Those who who believe in a both God and a Spirit or Life Force. Clearly such an approach cannot simultaneously address all interests in previous spirituality/well-being research. For example we do not combine our cognitive measure of belief with behavioral measures or attitudinal measures. However, this cost to scope is paid for with benefits to precision. The spiritual indicators measure cognitive states in which distinct varieties of skeptical beliefs, traditional religious beliefs, and non-traditional spirit beliefs are straightforwardly disentangled, and measured within people. Third, to understand how spirit beliefs affect growth, stability, or decline in psychological well-being over time, we leverage longitudinal responses from a nationally diverse probability sample year over nine years (2010 to 2018 New Zealand Attitudes and Values Study, NZAVS).

@@

Table 1 about here:

## Sample Demographics

@@

## Statistical Analyses

Statistical analysis was performed using R version 4.0.2 R version 3.6.1 (2020-06-22) 2019-07-05). We analyze data from participants who responded to the NZAVS survey at least three times between 2010 (Time 2) and 2018 (Time 10), resulting in a sample of  21,705 (N = 20,979 with complete responses).  To model growth in subjective well-being we use generalized linear mixed models. In our models, we account for the non-linear effects of time within participants by including the interaction of years (a continuous variable) with Beliefs (a four-level categorical indicator). The baseline for this indicator is Skepticism about a Spirit/Life Forces and Skepticism about a God. To handle the dependencies and heterogeneity introduced from the repeated measures, we include individual ID as an effect modelled as random.

## Potential Confounders and Missing Data

In a supplemental analysis we include standard demographic indicators used in New Zealand Attitudes and Values Research: [NZAVS modelling standards](https://cdn.auckland.ac.nz/assets/psych/about/our-research/nzavs/NZAVSTechnicalDocuments/NZAVS-Technical-Documents-e11-Default-Statistical-Standards.pdf). We centered and scaled the education, deprivation, and political conservativism. We centred age was centred at and put into decade-long intervals. The model equation(s) are described in Equation 2.

Because missing values can result in biased estimates [@Blackwell2017-oq], we multiply imputed 10 datasets using the AmeliaII package in R[@Honaker2011-yu].We conducted a parallel analysis averaging over the multiply imputed datasets, without any substantive difference to inferences. Additionally the results were robust to filtering of participants of N >3 to N = 9 waves. Additionally, the findings reported below were robust to modeling only the effects of years and beliefs, without adjusting for the standard NZAVS demographic covariates. The supplement describing these analysis is located at: https://github.com/jacopastorius/nzavs_spiritbeliefWellbeing.

### System and packages

The analysis was performed using R version 4.0.2 (2020-06-22). The Platform was x86_64-apple-darwin17.0 (64-bit) Running under: macOS Catalina 10.15.6
We are greatful to the contributors and mantainers of the following packages:
R [Version 4.0.2; @R-base] and the R-packages *Amelia* [Version 1.7.6; @R-Amelia], *citr* [Version 0.3.2; @R-citr], *dplyr* [Version 1.0.2; @R-dplyr], *equatiomatic* [Version 0.1.0.9000; @R-equatiomatic], *forcats* [Version 0.5.0; @R-forcats], *ggeffects* [Version 0.15.0; @R-ggeffects], *gghighlight* [Version 0.3.0; @R-gghighlight], *ggplot2* [Version 3.3.2; @R-ggplot2], *ggsci* [Version 2.9; @R-ggsci], *lme4* [Version 1.1.23; @R-lme4], *Matrix* [Version 1.2.18; @R-Matrix], *papaja* [Version 0.1.0.9997; @R-papaja], *parameters* [Version 0.8.6.1; @R-parameters], *patchwork* [Version 1.0.1; @R-patchwork], *prettycode* [Version 1.1.0; @R-prettycode], *purrr* [Version 0.3.4; @R-purrr], *Rcpp* [Version 1.0.5.1; @R-Rcpp_a; @R-Rcpp_b], *readr* [Version 1.3.1; @R-readr], *report* [Version 0.1.0; @R-report], *see* [Version 0.5.2.1; @R-see], *sjPlot* [Version 2.8.4; @R-sjPlot], *stringr* [Version 1.4.0; @R-stringr], *styler* [Version 1.3.2; @R-styler], *table1* [Version 1.2; @R-table1], *texreg* [Version 1.37.5; @R-texreg], *tibble* [Version 3.0.4; @R-tibble], *tidyr* [Version 1.1.2; @R-tidyr], and *tidyverse* [Version 1.3.0; @R-tidyverse]"

@@

## Supplement

@@

\newpage
# Supplement Model Equations

## Priors for the Personal Wellbeing model

Model priors for Personal Wellbeing including group-level ID intercept and beliefs slopes are as follows

$$
\begin{aligned}
\operatorname{Personal Well Being}_{ij} \sim \text{Normal}(\mu_{ij},\sigma^2) \nonumber\\
\mu_{ij}\sim \boldsymbol{\alpha +\beta} \nonumber\\
\boldsymbol{\alpha} \sim \alpha_{0} +\alpha_j \nonumber\\
\alpha_{0} \sim \text{StudentT}(3,7.2,2.5)\nonumber\\
\boldsymbol{\beta}\sim \boldsymbol{b} + \beta_j \nonumber\\
\boldsymbol{b}\sim\text{Normal}(0,2.5) \nonumber\\
\begin{bmatrix}
\alpha_j \\
\beta_j
\end{bmatrix}
\sim 
\text{MVNormal}
\begin{pmatrix}
\begin{bmatrix}
0\\
0
\end{bmatrix}
,\boldsymbol{S} \nonumber
\end{pmatrix}\\
\boldsymbol{S} = 
\begin{pmatrix}
0 & \sigma_{\beta_j} \nonumber
\end{pmatrix} \boldsymbol{R} \begin{pmatrix}
\sigma_{\alpha_j} & 0 \\
0 & \sigma_{\beta_j} \nonumber
\end{pmatrix}\\
\sigma_{\alpha_j} \sim 
\text{HalfCauchy}(0,2.5) \nonumber\\
\sigma_{\beta_j} \sim 
\text{HalfCauchy}(0,2.5) \nonumber\\
\boldsymbol{R}\sim \text{LKJcorr}(1) \nonumber\\
\sigma  \sim  \operatorname{HalfStudent T}^{+}(3, 0, 2.5) 
\end{aligned}
$$

\newpage
## Priors for the Life Satisfaction model

Model priors for Life Satisfaction Model with including group-level ID intercept and belief slopes are as follows:
$$
\begin{aligned}
\operatorname{Life Satisfaction}_{ij} \sim \text{Normal}(\mu_{ij},\sigma^2) \nonumber\\
\mu_{ij}\sim \boldsymbol{\alpha +\beta} \nonumber\\
\boldsymbol{\alpha} \sim \alpha_{0} +\alpha_j \nonumber\\
\alpha_{0} \sim \text{StudentT}(3,5.5, 2.5)\nonumber\\
\alpha_j \sim \text{StudentT}+(3,0,10)\nonumber\\
\boldsymbol{\beta}\sim \boldsymbol{b} + \beta_j \nonumber\\
\boldsymbol{b}\sim\text{Normal}(0,2.5) \nonumber\\
\begin{bmatrix}
\alpha_j \\
\beta_j
\end{bmatrix}
\sim 
\text{MVNormal}
\begin{pmatrix}
\begin{bmatrix}
0\\
0
\end{bmatrix}
,\boldsymbol{S} \nonumber
\end{pmatrix}\\
\boldsymbol{S} = 
\begin{pmatrix}
\sigma_{\alpha_j} & 0 \\
0 & \sigma_{\beta_j} \nonumber
\end{pmatrix} \boldsymbol{R} \begin{pmatrix}
\sigma_{\alpha_j} & 0 \\
0 & \sigma_{\beta_j} \nonumber
\end{pmatrix}\\
\sigma_{\alpha_j} \sim 
\text{HalfCauchy}(0,1) \nonumber\\
\sigma_{\beta_j} \sim 
\text{HalfCauchy}(0,1) \nonumber\\
\boldsymbol{R}\sim \text{LKJcorr}(1) \nonumber\\
\sigma  \sim  \operatorname{HalfStudent T}^{+}(3, 0, 2.5) 
\end{aligned}
$$
\newpage

## Deterministic model

The deterministic portion of both the Personal Wellbeing and the Life Satisfaction models is as follows: 
$$
\begin{aligned}
& \quad  \boldsymbol{b} = \\
& \quad \beta_{1}(\operatorname{Beliefs}_{\operatorname{SpiritExcludesGod}})\ + \\
&\quad \beta_{2}(\operatorname{Beliefs}_{\operatorname{GodAndSpirit}})\ +  \beta_{3}(\operatorname{Beliefs}_{\operatorname{GodExcludesSpirit}})\ +\\
& \quad \beta_{4}(\operatorname{Years} \times \operatorname{Beliefs}_{\operatorname{SpiritExcludesGod}})\ + \\ 
& \quad \beta_{4}(\operatorname{Years} \times \operatorname{Beliefs}_{\operatorname{GodAndSpirit}})\ + \\
&\quad \beta_{6}(\operatorname{Years} \times \operatorname{Beliefs}_{\operatorname{GodExcludesSpirit}}) \\ 
\end{aligned}
$$
\newpage
